High-Performance Primitives for Distributed Memory Architectures using Asynchronous Events

### 2012-11-11 doc/llr-pldi2013/impl.tex
updates

In addition to the RMA capabilities, GASNet provides {\em active messages} for 
inter-node communication.  Active messages consist of a command and payload that
are sent by one node to another without any previous coordination.  Upon arrival
at a destination node, a handler routine is invoked to process the message.
%The handler may issue a response to the sender, but this is optional, and avoided
%as much as possible in our implementation due to the latencies involved in waiting
%for a response.  
One example of a case where we use active messages involves the {\tt spawn} call on a {\tt Processor}
object.  If the processor is not on the same node as the requestor, the information
%about the task (e.g. the task ID and arguments, the actual target {\tt Processor},
%the prerequisite event (if any), and the completion event ID) 
for the task is sent in an
active message to the node containing the target processor.  The recipient node
unpacks the message, and either places the task on the ready queue for the target
processor or marks it as being deferred until the prerequisite event has triggered.
%No response is sent to the original requestor.  The requestor already knows the event
%that corresponds to the completion of this task (since it was provided in the active
%message) - there is no utility in knowing exactly when the active message itself has
%been processed.

@@

Events are created on demand and are {\em owned} by the node on which they
were created.  The creation of an event occurs with no inter-node communication.
The space of event
handles is divided across the nodes at start-up time, allowing each node to assign handles
to new events without the risk of conflicting with another node's assignments.  The static division
of event handles also permits any node to determine the owning node of an event
without any communication.
%No broadcast of
%the event creation is required either, as the static division is sufficient to allow any
%node to correctly determine an event's owning node when (and only if) an operation on that
%event is performed on the other node.

%At event creation time, the owning node allocates a data structure to track the state of the event
%(i.e. triggered or not) as well as record a list of the operations that are known to be dependent
%on the event (e.g. initiation of a copy operation, placing a task on a processor's ready queue, waking
%up a waiting task).  However, the owning node only keeps a list of the dependent operations from the same
%node.  Every other node also allocates a corresponding data structure (the first time the event
%is referenced) to remember if the event has already triggered and to record its local dependent operations.
%Arbitrarily many dependent operations from a
%single node are aggregated into a single {\em event subscription} active message that is sent to the
%owning node.  The owning node keeps a bitmask of which other nodes have subscribed to the event, and when the
%event finally does trigger, a single {\em event trigger} active message is sent to each subscribing node,
%which notes that the event has triggered (in case queries come after the trigger) and executes the list
%of local operations that were dependent on that event.  (In the case that the triggering of the event has
%occurred before the reception of an event subscription, a trigger message is sent immediatel to the 
%new subscriber.)
When a new event is created, the owning node allocates a data structure to track the state of
the event ({\em triggered} or {\em untriggered}) as well as to record a list of dependent operations
(e.g. copy operation or task launch) from the same node called {\em waiters}.
The first operation dependent on an event
from a remote node will allocate the same data structure on the remote node.  A single
{\em event subscription} active message is then sent from the remote node to the owning node indicating
that the remote node should be informed when the event triggers.  Arbitrarily many dependent operations on
the remote node can then be added to the list of waiters without any additional communication.
When the event does trigger, the owner node notifies all local waiters and
sends a single {\em event trigger} active message to each node from which it has seen an event subscription
to notify remote waiters.
In the case where an event triggers while a subscription message is in flight from a remote node, the owner node will
immediately respond with a trigger message.

The actual triggering of an event may occur on any node.  If it occurs on a node other than the owning
node, an {\em event trigger} active message is sent from the triggering node back to the owning node, which
then forwards that message to all the other subscribed nodes.  The triggering node will automatically
notify its waiters and no message is sent from the owning node back to the triggering node.
%with the exception of the triggering node, which was
%able to execute its locally dependent operations (if any) immediately.  
While a remote trigger of an event can result in
the latency of a triggering operation being at least two active message flight times, it bounds the number of active
messages required per event to $2N-2$ where $N$ is the number of nodes 
%(which can be much smaller than the number of dependent operations) 
monitoring the event.  An alternative
would have been to share the list of subscribers so that the triggering node could notify all interested
nodes directly.  However, such an algorithm is both more complicated due to race conditions, and requires $O(N^2)$
active messages.  Any algorithm that is super-linear with the node count will not scale well, 
and as we will see in Section~\ref{subsec:eventmicro}, the latency of a single event trigger
active message is very small. 
%and any algorithm that is super-linear with the node count will not scale well.
%either is or will soon be an scalability issue for large systems.

@@

Like events, deferred locks, or locks from here on, are created on demand by the node on which the creation request was made.  The lock
handles are also statically divided across the nodes, allowing the creation to be done without inter-node
communication.  Whereas event ownership is static, locks may migrate; the {\em creating}
node is the initial owner, but that ownership can be transferred to other nodes.

@@

When a lock request is made, an event is created to track when the grant occurs.  The current node then
examines its copy of the lock data structure to determine if it is the owner.
%(creating it if this the first operation for that lock on that node).  
If the current node is the owner and the lock isn't
held, the lock is granted immediately and the event is triggered.  If the lock is held, the event is added
to the list of local waiters.  Note that the event associated with
the lock request is the only data that must be stored.
%(This is the only thing that is stored.  The actual ``requestor'' of the lock
%is implicitly captured in which operations are dependent on the lock grant event.)  
If the current node isn't the owner, a {\em lock request} active message is
sent to the most recently known owner.  If the receiver of a lock request message is no
longer the owner, it forwards the message on to the node it has recorded as the owner.
%(If that information proves to be stale, the request is forwarded
%by the receiving node, eventually catching up to the current owner.)  
If the current owner's status shows
the lock is currently held, the bit for the requesting node is set in the remote waiters bitmask.  If the lock is
not held, the ownership of the lock is given to the requesting node via a {\em lock transfer} active
message.  A lock transfer message includes the bitmask of remote waiters and an up-to-date copy 
of the lock's payload.  The inclusion of the payload in the active message is the reason for the 
4KB size limit specified on payloads in Section~\ref{subsec:locks}.

>>>>

In addition to the RMA capabilities, GASNet provides {\em active messages} for 
inter-node communication.  Implementation of all runtime features requiring communication
rely heavily on active messages.  Active messages consist of a command and payload that
are sent by one node to another without any previous coordination.  Upon arrival
at a destination node, a handler routine is invoked to process the message.
%The handler may issue a response to the sender, but this is optional, and avoided
%as much as possible in our implementation due to the latencies involved in waiting
%for a response.  
%One example of a case where we use active messages involves the {\tt spawn} call on a {\tt Processor}
%object.  If the processor is not on the same node as the requestor, the information
%about the task (e.g. the task ID and arguments, the actual target {\tt Processor},
%the prerequisite event (if any), and the completion event ID) 
%for the task is sent in an
%active message to the node containing the target processor.  The recipient node
%unpacks the message, and either places the task on the ready queue for the target
%processor or marks it as being deferred until the prerequisite event has triggered.
%No response is sent to the original requestor.  The requestor already knows the event
%that corresponds to the completion of this task (since it was provided in the active
%message) - there is no utility in knowing exactly when the active message itself has
%been processed.

@@

Events are created on demand and are {\em owned} by the node on which they
are created.  Creating an event requires no inter-node communication.
The space of event
handles is divided across the nodes at start-up time, allowing each node to assign handles
to new events without conflicting with another node's assignments.  The static division
of event handles also permits any node to determine the owning node of an event
without communication.
%No broadcast of
%the event creation is required either, as the static division is sufficient to allow any
%node to correctly determine an event's owning node when (and only if) an operation on that
%event is performed on the other node.

%At event creation time, the owning node allocates a data structure to track the state of the event
%(i.e. triggered or not) as well as record a list of the operations that are known to be dependent
%on the event (e.g. initiation of a copy operation, placing a task on a processor's ready queue, waking
%up a waiting task).  However, the owning node only keeps a list of the dependent operations from the same
%node.  Every other node also allocates a corresponding data structure (the first time the event
%is referenced) to remember if the event has already triggered and to record its local dependent operations.
%Arbitrarily many dependent operations from a
%single node are aggregated into a single {\em event subscription} active message that is sent to the
%owning node.  The owning node keeps a bitmask of which other nodes have subscribed to the event, and when the
%event finally does trigger, a single {\em event trigger} active message is sent to each subscribing node,
%which notes that the event has triggered (in case queries come after the trigger) and executes the list
%of local operations that were dependent on that event.  (In the case that the triggering of the event has
%occurred before the reception of an event subscription, a trigger message is sent immediatel to the 
%new subscriber.)
When a new event is created, the owning node allocates a data structure to track the state of
the event ({\em triggered} or {\em untriggered}) as well as to record a list of dependent operations
(e.g. copy operation or task launch) from the same node called {\em waiters}.
The first operation dependent on an event
from a remote node allocates the same data structure on the remote node.  A single
{\em event subscription} active message is then sent from the remote node to the owning node indicating
that the remote node should be informed when the event triggers.  Arbitrarily many dependent operations on
the remote node can then be added to the list of waiters without any additional communication.
When the event does trigger, the owner node notifies all local waiters and
sends a single {\em event trigger} active message to each node from which it has seen an event subscription
to notify remote waiters.
In the case where an event triggers while a subscription message is in flight from a remote node, the owner node will
immediately respond with a trigger message.

The actual triggering of an event may occur on any node.  If it occurs on a node other than the owning
node, an {\em event trigger} active message is sent from the triggering node back to the owning node, which
then forwards that message to all the other subscribed nodes.  The triggering node will automatically
notify its waiters and no message is sent from the owning node back to the triggering node.
%with the exception of the triggering node, which was
%able to execute its locally dependent operations (if any) immediately.  
While a remote trigger of an event can result in
the latency of a triggering operation being at least two active message flight times, it bounds the number of active
messages required per event to $2N-2$ where $N$ is the number of nodes 
%(which can be much smaller than the number of dependent operations) 
monitoring the event.  An alternative
would have been to share the list of subscribers so that the triggering node could notify all interested
nodes directly.  However, such an algorithm is both more complicated (due to race conditions) and requires $O(N^2)$
active messages.  Any algorithm that is super-linear with the node count will not scale well, 
and as we will see in Section~\ref{subsec:eventmicro}, the latency of a single event trigger
active message is very small. 
%and any algorithm that is super-linear with the node count will not scale well.
%either is or will soon be an scalability issue for large systems.

@@

Like events, deferred locks (henceforth called simply locks) are
created on demand by the node of the creation request. 
Lock handles are also statically divided across the nodes,
allowing creation to be done without inter-node communication.
Whereas event ownership is static, locks may migrate; the {\em
  creating} node is the initial owner, but ownership can be
transferred to other nodes.

@@

When a lock request is made, an event is created to track when the grant occurs.  The current node then
examines its copy of the lock data structure to determine if it is the owner.
%(creating it if this the first operation for that lock on that node).  
If the current node is the owner and the lock isn't
held, the lock is granted immediately and the event is triggered.  If the lock is held, the event is added
to the list of local waiters.  Note that the event associated with
the lock request is the only data that must be stored.
%(This is the only thing that is stored.  The actual ``requestor'' of the lock
%is implicitly captured in which operations are dependent on the lock grant event.)  
If the current node isn't the owner, a {\em lock request} active message is
sent to the most recently known owner.  If the receiver of a lock request message is no
longer the owner, it forwards the message on to the node it has recorded as the owner.
%(If that information proves to be stale, the request is forwarded
%by the receiving node, eventually catching up to the current owner.)  
If the current owner's status shows
the lock is currently held, the bit for the requesting node is set in the remote waiters bitmask.  If the lock is
not held, the ownership of the lock is given to the requesting node via a {\em lock transfer} active
message, which includes the bitmask of remote waiters and an up-to-date copy 
of the lock's payload.  The inclusion of the payload in the active message is the reason for the 
4KB size limit specified on payloads in Section~\ref{subsec:locks}.

### 2012-11-11 doc/llr-pldi2013/interface.tex
updates

\section{Programming Interface}
\label{sec:interface}
The target set of clients for our interface are higher-level language
compilers and runtimes such as Legion\cite{Legion12} as well as 
advanced systems programmers.  This class of clients expect total control
over the underlying hardware and transparent performance from an
implementation.  To support these demands we have designed our interface
to be as low-level as possible.  In many cases we have chosen to trade-off
ease of programmability for performance.  We eschewed any features that would
automatically be performed without being directed by the client.
Our goal in the design of this interface is to provide a set of low-level,
high-performance primitives that are asynchronous and composable.

@@

\subsection{Events}
\label{subsec:events}
Events are the primary mechanism for describing dependences between operations in our system.
Lines 1-13 of Figure~\ref{fig:runtimeapi} show the interface for events.  An instance of the {\tt Event} type 
names a unique event in the system.  {\tt NO\_EVENT} (line 3)
is a special instance
of an event that by definition has always triggered.  The event interface
supports testing whether an event has triggered (line 5) and waiting on
an event to trigger (line 6).  While these methods can be useful, the preferred
use of events is passing them as preconditions for other operations.  The client can use the
{\tt merge\_events} call (line 7) to create an event that triggers only when every
event in a given set of events has triggered.  A code example using events follows in
Section~\ref{subsec:procs}.
%corresponding to the conjunction of a set of events.

@@

\subsection{Processors}
\label{subsec:procs}
Lines 14-24 of Figure~\ref{fig:runtimeapi} show the interface for {\tt Processor} objects which allow for
the creation of parallel computations.  Processors provide a way of naming 
every computational unit in the machine.  We describe how to discover processor handles
in Section~\ref{subsec:machmodel}.  In the case of our current interface 
processors are either individual CPU cores or discrete GPUs, but this is easily modified 
to include new processor types (line 20).   Processors support a single {\tt spawn}
method (line 23) that launches a new {\em task} on that processor.
Note that because the spawn operation
is invoked on a processor handle, it is possible to launch a task on any
processor from anywhere in the system.  
The spawn call takes an optional event that must trigger before the task can begin, and
always returns an event that will trigger when the task completes.

Figure~\ref{fig:procevents} shows a simple example using 
the processor and event interface.  The {\tt diamond\_task} launches four
sub-tasks with a diamond dependence pattern.
The {\tt merge\_events} call
merges the events {\tt eb} and {\tt ec} to create the shared dependence
for the last sub-task.

@@

An important difference between deferred locks and blocking locks is that the processor
that requests the lock doesn't have to be the one that uses it.  A common
convention in writing code with deferred locks is to acquire locks on behalf of a task being
launched.  Figure~\ref{fig:lockevents} illustrates this with a simple example.  The {\tt lock\_example}
function needs to launch two subtasks on two othe processors, each with exclusive access to some resource protected by a
lock named {\tt mutex}.  For each subtask, it first requests the lock on the subtaks behalf, making the
subtask wait until the lock has been acquired.  The parent task also requests that the runtime automatically
release the lock after each subtask has completed.  In the event dependency graph, solid arrows represent
the explicit events used in the code, while the dotted line represents the dynamic ordering that results
from the mutual exclusion guarantee of the lock.  (In this example, the second subtask is granted the lock first.)

When a lock is used to mediate access to a piece of data, the first thing a task will usually
do after being granted the lock is access the data.  To hide the latency of that
access as well, an additional feature of locks in our interface is the association of a lock with a
{\em payload}, a small (i.e. less than 4KB) 
piece of data that is moved around with the lock and guaranteed to be coherent while 
the lock is held.  The size of the payload is specified when the lock is created (line 35 of Figure~\ref{fig:runtimeapi}) and
the pointer to the local copy of the payload can be obtained from the {\tt payload\_ptr} method
(line 36).

Between deferred locks and barriers (described in Section~\ref{subsec:events}) clients 
have access to the same set of synchronization primitives that they have
in threading and bulk-synchronous interfaces.  However, these operations can now 
be composed with other asynchronous operations which is not possible in any other interface
of which we are aware.

@@

A physical region is an allocation of data in a single memory in the memory hierarchy.  Physical
regions are grouped into {\em classes} which are sets of physical regions that share the
same names (i.e. addresses) for elements, but don't necessarily share the same data layout.  This
allows pointers to elements within regions to be used, even if the regions are copied from one memory
to another.  Our
interface supports sub-classing, but the details are omitted due to space constraints.  The
interface allows copies between any pair of physical regions that are of
the same class (or between super- and sub-classes).

@@

Lines 39-61 of Figure~\ref{fig:runtimeapi} show a subset of the interface for physical regions.  
Each memory in the machine's memory hierarchy is named by a {\tt Memory} object (line 39).
The next object is a {\tt RegionMetaData} object (line 44), which defines a class of
regions.  Physical regions that belong to that class are created with the {\tt create\_instance}
method (line 51) and represented by {\tt RegionInstance} objects (line 56).  When creating a physical
region the client specifies the memory in which the physical region is going
to be allocated.  To maintain performance transparency, a new physical region can only be allocated
if the memory has sufficient capacity.
There is no automatic coherence of data between physical regions in the
same class.  It is the client's responsibility to manage data coherence using copies
between physical regions (line 59).  Like other operations copies can be
predicated on an event and return an event corresponding to completion.

If the only operations a task will perform on a region are reductions, a special reduction-only
instance can be created (line 52).  Rather than applying each reduction separately to the target instance,
the reduction operations can be accumulated in the reduction instance, and applied in bulk to 
the target using the {\em reduce\_to} operation (line 60).  This has several benefits.  First, it results
in much better performance, as we will show in section~\ref{subsec:reducmicro}), by reducing both
the total inter-node traffic and the overhead of that traffic while still allowing parallelism.
Second, a reduction-only instance is often smaller that the target instance, potentially improving
cache performance and helping address capacity issues for smaller memories (e.g. a GPU's device
memory).  For example, if reductions are being made to a single field in a larger structure, the 
reduction-only instance only needs to store that field and not the other fields that won't be 
modified by the reduction operation.

>>>>

\section{Programming Interface}
\label{sec:interface}
The clients for our interface are higher-level language
compilers and runtimes such as Legion\cite{Legion12} as well as 
advanced systems programmers.  These clients expect total control
over the underlying hardware and transparent performance from an
implementation.  To support these demands we have designed our interface
to be as low-level as possible.  In many cases we have chosen to trade-off
ease of programmability for performance.  We eschewed any features that would
automatically be performed without being directed by the client.
Our goal in the design of this interface is to provide a set of low-level,
high-performance primitives that are asynchronous and composable.

@@

\subsection{Events}
\label{subsec:events}
Events are the primary mechanism for describing dependences between operations in our system.
Lines 1-13 of Figure~\ref{fig:runtimeapi} show the interface for events.  An instance of the {\tt Event} type 
names a unique event in the system.  {\tt NO\_EVENT} (line 3)
is a special instance
of an event that by definition has always triggered.  The event interface
supports testing whether an event has triggered (line 5) and waiting on
an event to trigger (line 6), but the preferred
use of events is passing them as preconditions for other operations.  The client can use the
{\tt merge\_events} call (line 7) to create an event that triggers only when every
event in a given set of events has triggered.  A code example using events follows in
Section~\ref{subsec:procs}.
%corresponding to the conjunction of a set of events.

@@

\subsection{Processors}
\label{subsec:procs}
Lines 14-24 of Figure~\ref{fig:runtimeapi} show the interface for {\tt Processor} objects which allow for
the creation of parallel computations.  Processors provide a way of naming 
every computational unit in the machine.  We describe how to discover processor handles
in Section~\ref{subsec:machmodel}.  In the current interface 
processors are either individual CPU cores or discrete GPUs, but this is easily extended
to include new processor types (line 20).   Processors support a single {\tt spawn}
method (line 23) that launches a new {\em task} on that processor.
Note that because the spawn operation
is invoked on a processor handle, a task can be launched on a 
processor from anywhere in the system.  
The spawn call takes an optional event that must trigger before the task begins and
always returns an event that triggers when the task completes.

Figure~\ref{fig:procevents} shows an example using 
the processor and event interface.  The {\tt diamond\_task} launches four
sub-tasks with a diamond dependence pattern.
The {\tt merge\_events} call
merges the events {\tt eb} and {\tt ec} to create the shared dependence
for the last sub-task.

@@

An important difference between deferred locks and blocking locks is that the processor
that requests the lock doesn't have to be the one that uses it.  A common
convention in writing code with deferred locks is to acquire locks on behalf of a task being
launched.  Figure~\ref{fig:lockevents} illustrates this with a simple example.  The {\tt lock\_example}
function needs to launch two subtasks on two other processors, each with exclusive access to some resource protected by a
lock named {\tt mutex}.  For each subtask, it first requests the lock on the subtask's behalf, making the
subtask wait until the lock has been acquired.  The parent task also requests that the runtime automatically
release the lock after each subtask has completed.  In the event dependency graph, solid arrows represent
the explicit events used in the code, while the dotted line represents the dynamic ordering that results
from the mutual exclusion guarantee of the lock.  (In this example, the second subtask is granted the lock first.)

When a lock is used to mediate access to a piece of data, the first thing a task will usually
do after being granted the lock is access the data.  To hide the latency of that
access as well, an additional feature of locks in our interface is the association of a lock with a
{\em payload}, a small (i.e. less than 4KB) 
piece of data that is moved with the lock and guaranteed to be coherent while 
the lock is held.  The size of the payload is specified when the lock is created (line 35 of Figure~\ref{fig:runtimeapi}) and
the pointer to the local copy of the payload can be obtained from the {\tt payload\_ptr} method
(line 36).

With deferred locks and barriers (see Section~\ref{subsec:events}) clients 
have the same synchronization primitives as
in threading and bulk-synchronous interfaces.  However, these operations can now 
be composed with other asynchronous operations which is not possible in any other interface
of which we are aware.

@@

A physical region is an allocation of data in a single memory in the memory hierarchy.  Physical
regions are grouped into {\em classes}, which are sets of physical regions sharing the
same names (i.e. addresses) for elements, but not necessarily sharing the same data layout.  This
allows pointers to elements within regions to be used, even if the regions are copied from one memory
to another.  Our
interface supports sub-classing, but the details are omitted due to space constraints.  The
interface allows copies between physical regions of
the same class (or between super- and sub-classes).

@@

Lines 39-61 of Figure~\ref{fig:runtimeapi} show a subset of the interface for physical regions.  
Each memory in the machine's memory hierarchy is named by a {\tt Memory} object (line 39).
The next object is a {\tt RegionMetaData} object (line 44), which defines a class of
regions.  Physical regions that belong to that class are created with the {\tt create\_instance}
method (line 51) and represented by {\tt RegionInstance} objects (line 56).  When creating a physical
region the client specifies the memory in which the physical region is
to be allocated.  To maintain performance transparency, a new physical region can only be allocated
if the memory has sufficient capacity.
There is no automatic coherence of data between physical regions in the
same class.  It is the client's responsibility to manage data coherence using copies
between physical regions (line 59).  Like other operations copies can be
predicated on an event and return an event corresponding to completion.

If the only operations a task will perform on a region are reductions, a special reduction-only
instance can be created (line 52).  Rather than applying each reduction separately to the target instance,
the reduction operations can be accumulated in the reduction instance, and applied in bulk to 
the target using the {\em reduce\_to} operation (line 60).  This has several benefits.  First, it results
in much better performance, as we show in Section~\ref{subsec:reducmicro}, by reducing both
the total inter-node traffic and the overhead of that traffic while still allowing parallelism.
Second, a reduction-only instance is often smaller that the target instance, potentially improving
cache performance and helping address capacity issues for smaller memories (e.g. a GPU's device
memory).  For example, if reductions are being made to a single field in a larger structure, the 
reduction-only instance only needs to store that field and not the other fields that won't be 
modified by the reduction operation.
